Containers and R---the Rockerverse and beyond

### 2020-01-13 manuscript.Rmd
add first version of conclusions
closes #31 by mentioning rkt as potential future work

- what are next steps?
- support more complex set-ups with ready-to-use stacks, e.g., GPU, AI, ML
- Missing pieces?
- consolidation (e.g., via packages using `dockerfiler` and `stevedore`)
- Common themes
  - reproducibility
- will knowledge about containers continue to spread?
- what is needed for even more containers with R?
- the ability to move processing between services easily (e.g locally, one cloud providers VM, another cloud provider's Container-as-a-Service)
- running RStudio in a container is a common approach for reproducibility - _is there a need to run/control containers from a UI integrated in RStudio?_
- moving away from Docker focus towards containers?
  - is there a need for a wrapper for [podman's Varlink-API](https://github.com/containers/libpod/blob/master/API.md) (similar to stevedore) and buildah to manage containers and images respectively with alternative software stacks?
  - will there be control packages for Singularity support?
- sustain the overview and structure in this article in form of a CRAN Task View

>>>>

This article is a snapshot of the R-corner in a universe of applications built many-faced piece of software, Docker.
Docker became synonymous with applying the concept of containerisation to solve many challenges, such as reproducible environments in research, scalable deployments in heterogenous organisations, and the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
These are probably the thre common themes described in this work.

The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with continuous evolving landscape.
The use cases contributed by co-authors also have a degree of overlap.
A common language and understanding of good practices is still taking shape.
The easiness with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs currently leads to some fragmentation.
It is unclear if Docker alone as a core technology is really a feasible level of collaboration and abstraction.
Furthermore fragmentation may not be a bad sign but instead a reflection of a growing market, which is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.

However, at least _on the level of R packages_, some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g. `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
In this consolidation, the Rocker Project could play the role of a unifying and coordinating entity in the future.
Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
Currently containerisation happens more in the background at the system level.

New features, which make complex workflows accessible and reproducible, and variety in packages, even when they have overlapping features, are a signal and a support for a growing user base.
This growth is possibly the the most important goal for the foreseeable future in the _Rockerverse_.
The dominant position of Docker is a blessing and a curse for these goals.
It could be wise to start experimenting with non-Docker containerisation tools now, e.g. R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
A concrete effort by this work towards these pieces is to sustain the structure and captured status quo in a CRAN Task View _Containerization_.
